---
title: "Exercise 2, Ling Stats S2023"
subtitle: "Due Tuesday, February 7, 2023 via Canvas"
author: "Clio Begnal"
date: "February 7, 2023"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=TRUE, include=TRUE, results="hold", comment=NA)
library(tidyverse)
library(ggplot2)
```

# Getting set up.

-   Replace my name with your name in the title box above, and add the current date.
-   Within the `setup` code chunk above, load the tidyverse package.
-   Run the following code to read in the "I_jean" file from Joe Fruehwald's R tutorial.

```{r read in dataset}

jean <- read.delim("http://bit.ly/avml_ggplot2_data")
jean

```

# Orient yourself to the dataframe.

Answer the two questions below about the dataframe you've just read in. For these first two questions, you may show your code and/or simply answer them in prose.

1.  How many observations are in this dataframe? What does each observation seem to correspond to?

-   Insert your answer here.

```{r}
count(jean)
# Each observation is Jean speaking a word I in different contexts
```

2.  How many unique speakers are there in this dataset, and how many unique words?

-   Insert your answer here. **1 speaker - Jean 5 different words all with I vowel**

# Summarize the data.

Be sure to show your code for questions 3-7.

3.  What is the overall average vowel duration in this dataset?

```{r calculate overall vowel average}
# Your code here.
jean %>%
  summarize(mean_dur = mean(Dur_msec))
```

4.  For each unique word context, calculate the average vowel duration, average F1, and average F2.

```{r average dur, F1, F2 by word}
# Your code here.
jean %>% 
  group_by(Word) %>%
  summarize(mean_dur = mean(Dur_msec), mean_F1 = mean(F1), mean_F2 = mean(F2))
```

5.  Restricting yourself to cases where the word was "I" and the following segment was either a voiced or a voiceless stop, calculate the average vowel duration, average F1, and average F2.

```{r average dur, F1, F2 for "I"}
# Your code here.
jean %>%
  filter(Word == 'I') %>%
  filter(FolSegTrans %in% c('B','D','G','P','T','K')) %>%
  summarize(avg_dur = mean(Dur_msec), mean_F1 = mean(F1), mean_F2 = mean(F2))
```

6.  Using the same restricted dataset as in (5) above, create a new variable that codes for whether the final segment was a voiced or voiceless stop. Then calculate the average duration as a function of this variable.

```{r average vowel dur for voiced vs. voiceless context}
# Your code here.
jean %>%
  filter(Word == 'I') %>%
  filter(FolSegTrans %in% c('B','D','G','P','T','K')) %>%
  mutate(stops = ifelse(FolSegTrans %in% c('B','D','G'), 'voiced', 'voiceless')) %>%
  group_by(stops) %>%
  summarize(avg_dur = mean(Dur_msec))
  
```

7.  Using the `filter()`, `group_by()`, `count()`, and `arrange()` functions, make a table of counts that shows how many observations there are for each segment that can follow the word "I", ordered from most to least frequent.

```{r count following segments for the word "I"}
# Be sure to arrange your table from most to least frequent following segments.
jean %>%
  filter(Word=='I') %>%
  group_by(FolSegTrans) %>%
  count() %>%
  arrange(desc(n))
```

# Practice plotting.

Be sure to use the full "jean" dataset in this section (not the data summaries you created above).

8.  Make a plot that compares vowel durations for the different words in this dataset.

-   Which word has the longest vowel duration, on average? **I've does!**

```{r plot vowel dur by word}
# Your code here.
ggplot(jean, aes(Word, Dur_msec)) +
  geom_boxplot(fill = "lightblue", color = "blue")
```

9.  Is there a relationship between vowel duration and F1 in this dataset? Be sure to add a trend line to help any patterns jump out.

```{r plot vowel dur vs. F1}
# Your code here.
ggplot(jean, aes(F1, Dur_msec)) +
  geom_point() +
  stat_smooth(method = "lm")
```

**Seems like there's a positive relationship!**

10. Is there a relationship between vowel duration and F2 in this dataset? Be sure to add a trend line again.

```{r plot vowel dur vs. F2}
# Your code here.
ggplot(jean, aes(F2, Dur_msec)) +
  geom_point() +
  stat_smooth(method = "lm")
```

**Seems like there's a negative relationship!**

11. Restricting yourself to cases where the following consonant is either a voiced or voiceless stop, add a variable that codes for this (voiced vs. voiceless stop), then make a plot comparing vowel durations for these two contexts.

```{r plot vowel dur before voiced vs. voiceless stops}
# Your code here.
jean %>%
  filter(FolSegTrans %in% c('B','D','G','P','T','K')) %>%
  mutate(stops = ifelse(FolSegTrans %in% c('B','D','G'), 'voiced', 'voiceless')) %>%
  group_by(stops) %>%
  ggplot(aes(stops, Dur_msec))+
  geom_boxplot(fill = "lightblue", color = "blue")
```

12. About how long did this assignment take you, in total?

-   Your answer here. **Around 30-45 minutes!**
