---
title: "Exercise 3, Ling Stats S2023"
subtitle: "Due Friday, February 17, 2023 via Canvas"
author: "Clio Begnal"
date: "(17 February 2023)"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=TRUE, include=TRUE, comment=NA)
library(tidyverse)
library(Rling)
library(modeest)
data("ldt")
```

# Getting set up.
* Replace my name with your name in the title box above, and add the current date.
* Within the `setup` code chunk above, load the `tidyverse` package, the `Rling` package, and the `ldt` dataset within the `Rling` package.


# Orient yourself to the dataframe.
For this first question, you may show your code and/or simply answer it in prose.

1. How many observations are in this dataframe? What does each observation seem to correspond to?
```{r}
count(ldt)
```


* Insert your answer here.
**There are 100 observations and they represent the length, frequency, and mean reaction time of different words.**

# Summarize and explore the data.
Be sure to show your code in this section!

2a. What are the mean, median and mode values for the `Length` variable in this dataset?

```{r calculate the mean, median, and mode for Length}
# Your code here:
mean(ldt$Length)

median(ldt$Length)

ldt %>%
  group_by(Length) %>%
  count() %>%
  arrange(desc(n))
```
**The mean is 8.23, the median is 8, and the mode is bimodal at 8 and 10.**

2b. Create a histogram and a q-q plot showing the distribution of the `Length` variable. Does `Length` appear relatively normally distributed?
  
  * Your answer here.

```{r plots for Length}
# Your code here:
qqnorm(ldt$Length)
qqline(ldt$Length)

ggplot(ldt, aes(Length)) +
  geom_histogram(bins=10,color="darkblue", fill="lightblue") +
  geom_density(aes(y=1.1*..count..))
```
**Length does  seem to be normally distributed as the histogram is close to a bell curve and the q-q plot looks somewhat linear.**

2c. Run a Shapiro-Wilk normality test for the `Length` variable. What does the result mean, and does it align with your prediction based on the plots?

  * Your answer here.

```{r Shapiro-Wilk test for Length}
# Your code here:
shapiro.test(sample(ldt$Length))
```
**The result here means that the result is from a normally distributed set of data as the p value is above 0.05.**

3a. What are the mean, median and mode values for the `Freq` variable in this dataset?

```{r calculate the mean, median, and mode for Freq}
# Your code here:
mean(ldt$Freq)

median(ldt$Freq)

ldt %>%
  group_by(Freq) %>%
  count() %>%
  arrange(desc(n))
```
**The mean is 3350.3`, the median is 310.5, and the mode is 0.**

3b. Create a histogram and a q-q plot showing the distribution of the `Freq` variable. Does `Freq` appear relatively normally distributed?

  * Your answer here.

```{r plots for Freq}
# Your code here:
qqnorm(ldt$Freq)
qqline(ldt$Freq)

ggplot(ldt, aes(Freq)) +
  geom_histogram(bins=10,color="darkblue", fill="lightblue") +
  geom_density(aes(y=1.1*..count..))
```
**Based on the graphs, Frequency does not seem to be normal at all.**

3c. Run a Shapiro-Wilk normality test for the `Freq` variable. What does the result mean, and does it align with your prediction based on the plots?

  * Your answer here.

```{r Shapiro-Wilk test for Freq}
# Your code here:
shapiro.test(sample(ldt$Freq))
```
**As the result is under 0.05 and is very small, it is highly unlikely that it is normal which agrees with my prediction.** 

4a. What are the mean, median and mode values for the `Mean_RT` variable in this dataset?

```{r calculate the mean, median, and mode for Mean_RT}
# Your code here:
mean(ldt$Mean_RT)

median(ldt$Mean_RT)

ldt %>%
  group_by(Mean_RT) %>%
  count() %>%
  arrange(desc(n))

mlv(ldt$Mean_RT)
```
**The mean is 808.25, the median is 784.94, and the mode is 789.00 according to the mlv command, but all the values actually occur once.**

4b. Create a density plot and a q-q plot showing the distribution of the `Mean_RT` variable. Does `Mean_RT` appear relatively normally distributed?

  * Your answer here.

```{r plots for Mean_RT}
# Your code here:
qqnorm(ldt$Mean_RT)
qqline(ldt$Mean_RT)

ggplot(ldt, aes(Mean_RT)) +
  geom_histogram(bins=10,color="darkblue", fill="lightblue") +
  geom_density(aes(y=1.1*..count..))
```
**No, is does not look normally distributed.**

4c. Run a Shapiro-Wilk normality test for the `Mean_RT` variable. What does the result mean, and does it align with your prediction based on the plots?

  * Your answer here.

```{r Shapiro-Wilk test for Mean_RT}
# Your code here:
shapiro.test(sample(ldt$Mean_RT))
```
**As the result is under 0.05, it is unlikely that it is normal which aligns with my prediction.**

# Thinking about normality.
6. What does the difference between the mean and median values for the `Mean_RT` variable tell you about the shape of this distribution?

* Your answer here, in prose.

7. Make a scatterplot that shows the relationship between the `Freq` and `Mean_RT` variables, and add a trend line. Is the trend line useful in this case? (i.e. Does it tell you something meaningful about the relationship between `Freq` and `Mean_RT` in this dataset?) Why or why not?

  * Your answer here, in prose. **No, the trend line and plot are very scattered and do not inform me of much. I cannot see a relationship, and the trendline is showing a very slightly negative relationship but it is hard to actually see because it is not a straight line.**

```{r make a scatterplot for Freq vs. Mean_RT}
# Your code here:
ggplot(ldt, aes(Freq,Mean_RT)) +
  geom_point()  +
  stat_smooth()
```

8. One way to deal with issues of non-normality is to remove outliers. Using the `mutate()` function, add the following columns to the `ldt` dataset:

  * A column showing the overall mean for `Freq` (every row should have the same value!).
  * A column showing the overall standard deviation for `ldt` (ditto).
  * A column where you take the mean and add 2 times the standard deviation (ditto).
  * Once you've added these columns, show how you can use the `filter()` function to remove words whose frequency is more than 2 times the standard deviation away from the mean.

```{r remove words with Freq > 2 SDs from the mean}
# Your code here:
ldt %>%
  mutate(overall_mean=mean(Freq), overall_SD=(sd(Freq)), mean_plus_2SD=(mean(Freq)+(2*sd(Freq)))) %>%
  filter(Freq < mean_plus_2SD) 

```


9. About how long did this assignment take you, in total?

* Your answer here. **Around 45 minutes!**
