---
title: "Exercise 4, Ling Stats S2023"
subtitle: "Due Friday, February 24, 2023 via Canvas"
author: "Clio Begnal"
date: "24 February 2023"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=TRUE, include=TRUE, comment=NA)
library(tidyverse)
library(languageR)
data('etymology')
?etymology
```

# Getting set up
* Replace my name with your name in the title box above, and add the current date.
* Within the `setup` code chunk above, load the `tidyverse` package.
* If you haven't done so before, you will need to install and then load the `languageR` package. Then load the `etymology` dataset.
* Open up the documentation (i.e. the help file) for `etymology` to read about the variables in this dataset.

**This assignment is more open-ended than previous assignments.
The goal is for you to use the tools you have acquired so far to explore
the Dutch `etymology` data set from the `languageR` package.**


# Getting oriented to the data

1. How many observations are in the etymology data set, and what does each observation consist of?
```{r}
head(etymology)
count(etymology)
```


* Your answer here, in prose. **There are 285 observations in this data set, and they refer to 14 variables: verbs, written frequency, numeric vector count, mean log bigram frequency, inflectional entropy, auxiliary levels, regularity, length, denominative, age, valency, and written spoken ratio.**

2. For the variable `EtymAge`, what are the possible levels of this factor? What do you think this variable is coding for?

```{r}
levels(etymology$EtymAge)
```

* Your answer here, in prose. **The factor EtymAge is referring to the age of the verb, and if it comes from Dutch, Dutch German, West Germanic, Germanic, or Indo-European origin.**

3. Using some knowledge about historical linguistics (hopefully), what order do you think the levels for "EtymAge" should be in? Re-order the levels so that the ordering makes more sense.

```{r reorder the levels of EtymAge}
# Your code here
levels(etymology$EtymAge) <- c("IndoEuropean", "Germanic", "WestGermanic","DutchGerman","Dutch")
levels(etymology$EtymAge)
```

# Explore the data set

4. Still working with `EtymAge`, how many observations are there for each level? What proportion of observations fall into each category?

```{r describe properties of EtymAge}
# Your code here
etymology %>%
  group_by(EtymAge) %>%
  summarize(n=n()) %>%
  mutate(prop=(n/sum(n)))
```

5. Now choose a categorical variable other than `EtymAge` and describe it:
What are the possible levels, and how many observations are there for each level? What proportion of observations fall into each category?

```{r choose a categorical variable and describe its properties}
# Your code here
levels(etymology$Regularity)

etymology %>%
  group_by(Regularity) %>%
  summarize(n=n()) %>%
  mutate(prop=(n/sum(n)))
```

6. Choose one more categorical variable and describe it, as well: What are the possible levels, and how many observations are there for each level? What proportion of observations fall into each category?

```{r choose another categorical variable and describe its properties}
levels(etymology$Auxiliary)

etymology %>%
  group_by(Auxiliary) %>%
  summarize(n=n()) %>%
  mutate(prop=(n/sum(n)))
```

7. Now choose one of the categorical variables you described above and make three plots with it:

* One showing the counts for the different levels using stacked bars,
* One showing the proportions of the different levels,
* And one showing the counts for the different levels using *un*stacked bars.

You may use the following code to create a "dummy variable" for plotting purposes, if you like:

```{r plot your variable three ways, eval=F}
etymology$DummyVar <- rep("AllVerbs", length(etymology$Verb))
etymology$DummyVar <- as.factor(etymology$DummyVar)

ggplot(etymology, aes(DummyVar, fill=Regularity)) +
  geom_bar(color = 'black') +
  xlab("DummyVar") +
  ylab("Frequency Count")

ggplot(etymology, aes(DummyVar, fill=Regularity)) +
  geom_bar(position = "fill", color = "black") +
  xlab("DummyVar") +
  ylab("Proportion")

ggplot(etymology, aes(DummyVar, fill=Regularity)) +
  geom_bar(position = "dodge", color = "black") +
  xlab("DummyVar") +
  ylab("Frequency Count")

```

8. Now choose two of the categorical variables you described above and investigate the relationship between them: Provide tables of counts and proportions, along with at least two graphs. Also provide a brief prose description of your findings.

* Your prose description here.
**The regularity of verbs is very close to being 50/50, and West Germanic verbs are the most frequent.**

```{r examine two categorical variables}
# Your code here
ggplot(etymology, aes(EtymAge, fill=Regularity)) +
  geom_bar(color = 'black') +
  xlab("EtymAge") +
  ylab("Frequency Count")

ggplot(etymology, aes(EtymAge, fill=Regularity)) +
  geom_bar(position = "fill", color = "black") +
  xlab("EtymAge") +
  ylab("Proportion")

ggplot(etymology, aes(EtymAge, fill=Regularity)) +
  geom_bar(position = "dodge", color = "black") +
  xlab("EtymAge") +
  ylab("Frequency Count")
```

9. Repeat the exercise above with a different pair of variables. Choose another pair of categorical variables and investigate the relationship between them: Provide tables of counts and proportions, along with at least two graphs. Also provide a brief prose description of your findings.

* Your prose description here.
**I found that hebben verbs have the highest count in West Germanic age, and zijnheb verbs are fairly recent in age.** 
```{r examine two more categorical variables}
# Your code here
ggplot(etymology, aes(EtymAge, fill=Auxiliary)) +
  geom_bar(color = 'black') +
  xlab("EtymAge") +
  ylab("Frequency Count")

ggplot(etymology, aes(EtymAge, fill=Auxiliary)) +
  geom_bar(position = "fill", color = "black") +
  xlab("EtymAge") +
  ylab("Proportion")

ggplot(etymology, aes(EtymAge, fill=Auxiliary)) +
  geom_bar(position = "dodge", color = "black") +
  xlab("EtymAge") +
  ylab("Frequency Count")
```

10. Next, choose a *numerical* variable, and investigate its relationship to one of the categorical variables. Provide a few descriptive statistics (such as means, medians, and/or standard deviations), along with at least one graph. Also provide a brief prose description of your findings.

* Your prose description here. 
**I found that the mean and median are close to 8, and the standard deviation is 1.67. West Germanic verbs seem to have the largest counts.**

```{r examine a numerical and a categorical variable}
# Your code here
mean(etymology$WrittenFrequency)
median(etymology$WrittenFrequency)
sd(etymology$WrittenFrequency)

ggplot(etymology, aes(WrittenFrequency, fill=EtymAge)) +
  geom_histogram(bins=10, color='black')
```

11. Now, imagine that you want to re-code the `LengthInLetters` variable, so that any words with a length of 4 or less are now in a category called "short", and words with a length of 5 or more are in a category called "long".

First show how you could do that, and then give descriptive statistics (counts and proportions) comparing the number of short vs. long words for verbs of different etymological ages.

Also include a graph comparing either the number or proportion of short vs. long words for different etymological ages.

```{r recode and then describe LengthInLetters}
# Your code here
etymology2 <- mutate(etymology, lengths = (ifelse(LengthInLetters > 4, 'long', 'short'))) 
etymology$LengthInLetters <- etymology2$lengths

ggplot(etymology, aes(EtymAge, fill=LengthInLetters))+
  geom_bar(position='fill', color='black') +
  xlab('Etymological Age')+
  ylab('Proportions')
```
```{r}
etymology %>%
group_by(LengthInLetters,EtymAge) %>%
  summarize(n=n()) %>%
  mutate(props=(n/sum(n)))
```

12. Finally, describe how you decided to format your plot: What variable is on the x-axis, and why? Did you use color ("fill") in your plot, and why? Do you think it's better to plot counts vs. proportions in this case, and why?

* Your answers here, in prose.
**I used Etymological age on the x-axis because it is not numerical, whereas the count is numerical and used on the y-axis. I think it's better to plot proportions because you can see which etymological ages have mainly long or short verbs.**